---
title: "DLPFC"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

# Introduction
Here, we apply KODAMA to analyze the human dorsolateral prefrontal cortex (DLPFC) data by 10x Visium from [Maynard et al., 2021](https://www.nature.com/articles/s41593-020-00787-0). The links to download the raw data and H&E full resolution images can be found in the [LieberInstitute/spatialLIBD](https://github.com/LieberInstitute/spatialLIBD) github page.

# Loading the required libraries
```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
library("nnSVG")
library("scater")
library("scran")
library("scry")
library("SPARK")
library("harmony")
library("Seurat")
library("spatialLIBD")
library("KODAMAextra")
library("mclust")
library("slingshot")
library("irlba")
library("Rnanoflann")
library("ggpubr")

```


# Download the dataset
```{r, eval=FALSE}
spe <- fetch_data(type = 'spe')
```

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
load("../DLFPC.RData")

 
plot_slide =  function (xy, slide, labels, col = NULL, nrow = 1, scales = "free",size.dot = 1,
                        size.legend.text=10,size.legend.title=10,size.legend.dot=5,size.strip.text=10)
{
  if (is.null(col)) {
    labels = as.factor(labels)
    nn = length(levels(labels))
    col = rainbow(nn)
  }
  df <- data.frame(xy, slide, labels)
  colnames(df) = c("x", "y", "slide", "labels")
  df$slide = as.factor(df$slide)
  df$labels = as.factor(df$labels)
  ggplot(df, aes(x, y, color = labels)) + geom_point(size = size.dot) +
    facet_wrap(~slide, nrow = nrow, scales = scales) + theme_bw() +
    theme(legend.position = "bottom", axis.title = element_blank(),
          legend.text = element_text(size = size.legend.text),  
          legend.title = element_text(size = size.legend.title), 
          strip.text = element_text(size = size.strip.text, face = "bold"), 
          axis.text = element_blank(), axis.ticks = element_blank(),
          panel.grid = element_blank()) +
    scale_color_manual("Domain",  values = col, drop = FALSE) + guides(color = guide_legend(nrow = 1,  override.aes = list(size = size.legend.dot)))  # Legend
}
plot_slide_minimal <- function(xy, slide, labels, col = NULL, nrow = 1,
                               size.dot = 3, size.legend.dot = 5) {
  if (is.null(col)) {
    labels <- as.factor(labels)
    nn <- length(levels(labels))
    col <- rainbow(nn)
  }

  df <- data.frame(xy, slide, labels)
  colnames(df) <- c("x", "y", "slide", "labels")
  df$slide <- as.factor(df$slide)
  df$labels <- as.factor(df$labels)

  ggplot(df, aes(x, y, color = labels)) +
    geom_point(size = size.dot) +
    facet_wrap(~slide, nrow = nrow) +
    theme_void() +
    theme(
      legend.position = "none",
      strip.background = element_blank(),
      strip.text = element_blank(),
      panel.spacing = unit(0, "lines")  # 👈 Make facets closer
    ) +
    scale_color_manual(values = col)
}
```

## Extract the metadata information
```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
n.cores=40
splitting = 100
spatial.resolution = 0.3
aa_noise=3
gene_number=2000
graph = 20
seed=543210


set.seed(seed)
ID=unlist(lapply(strsplit(rownames(colData(spe)),"-"),function(x) x[1]))
samples=colData(spe)$sample_id
rownames(colData(spe))=paste(ID,samples,sep="-")

txtfile=paste(splitting,spatial.resolution,aa_noise,2,gene_number,sep="_")

sample_names=c("151507",
               "151508",
               "151509",
               "151510",
               "151669",
               "151670",
               "151671",
               "151672",
               "151673",
               "151674",
               "151675",
               "151676")
subject_names= c("Br5292","Br5595", "Br8100")
metaData = SingleCellExperiment::colData(spe)
expr = SingleCellExperiment::counts(spe)
sample_names <- paste0("sample_", unique(colData(spe)$sample_id))
sample_names <-  unique(colData(spe)$sample_id)
dim(spe)
# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
table(is_mito)

# calculate per-spot QC metrics
spe <- addPerCellQC(spe, subsets = list(mito = is_mito))

# select QC thresholds
qc_lib_size <- colData(spe)$sum < 500
qc_detected <- colData(spe)$detected < 250
qc_mito <- colData(spe)$subsets_mito_percent > 30
qc_cell_count <- colData(spe)$cell_count > 12

# spots to discard
discard <- qc_lib_size | qc_detected | qc_mito | qc_cell_count
table(discard)
colData(spe)$discard <- discard
# filter low-quality spots
spe <- spe[, !colData(spe)$discard]
dim(spe)


spe <- filter_genes(
  spe,
  filter_genes_ncounts = 2,   #ncounts
  filter_genes_pcspots = 0.5,
  filter_mito = TRUE
)

dim(spe)

sel= !is.na(colData(spe)$layer_guess_reordered)
spe = spe[,sel]
dim(spe)

spe <- computeLibraryFactors(spe)
spe <- logNormCounts(spe)

 subjects=colData(spe)$subject
 labels=as.factor(colData(spe)$layer_guess_reordered)
 xy=as.matrix(spatialCoords(spe))
 samples=colData(spe)$sample_id
 
 cols_cluster <- c("#0000b6", "#81b29a", "#f2cc8f","#e07a5f",
          "#cc00b6", "#81ccff", "#33b233")

  plot_slide(xy,samples,labels,col=cols_cluster,size.dot = 1)
  
```

```{r, fig.width=14, fig.height=3, warning=FALSE, message=FALSE, echo=FALSE}
png("output/Figures/DLPFC/Fig 3a.png",width = 12000,height = 1100,bg = "transparent")
plot_slide_minimal(xy,samples,labels,col=cols_cluster,size.dot = 5)
dev.off()
```

## Gene selection
The identification of genes that display spatial expression patterns is performed using the SPARKX method ([Zhu et al. (2021)](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02404-0)). The genes are ranked based on the median value of the logarithm value of the p-value obtained in each slide individually.


```{r, fig.width=10, fig.height=4,results = "hide"}


top=multi_SPARKX(spe,n.cores=n.cores)

data=as.matrix(t(logcounts(spe)[top[1:gene_number],]))

genes=spe@rowRanges@elementMetadata$gene_name
names(genes)=spe@rowRanges@elementMetadata$gene_id

samples=colData(spe)$sample_id
labels=as.factor(colData(spe)$layer_guess_reordered)
names(labels)=rownames(colData(spe))
subjects=colData(spe)$subject
genes=rowData(spe)[,"gene_name"]
names(genes)=rowData(spe)$gene_id
genes_top=genes[top[1:gene_number]]


```


# Patient Br5595


```{r, fig.width=10, fig.height=4}

subject_names="Br5595"
nclusters=5

spe_sub <- spe[, colData(spe)$subject ==  subject_names]
 # subjects=colData(spe_sub)$subject
dim(spe_sub)
#  spe_sub <- runPCA(spe_sub, 50,subset_row = top[1:gene_number], scale=TRUE)
#pca=reducedDim(spe_sub,type = "PCA")[,1:50]
  
spe_sub <- spe[, colData(spe)$subject ==  subject_names]
sel= subjects ==  subject_names
        
data_Br5595=data[sel,top[1:gene_number]]
        
RNA.scaled=scale(data_Br5595)
pca_results <- irlba(A = RNA.scaled, nv = 50)
pca_Br5595 <- pca_results$u %*% diag(pca_results$d)[,1:50]
rownames(pca_Br5595)=rownames(data_Br5595)
colnames(pca_Br5595)=paste("PC",1:50,sep="")
labels=as.factor(colData(spe_sub)$layer_guess_reordered)
names(labels)=rownames(colData(spe_sub))
xy=as.matrix(spatialCoords(spe_sub))
rownames(xy)=rownames(colData(spe_sub))
samples=colData(spe_sub)$sample_id

        subject_names_Br5595=colData(spe_sub)$subject

plot(pca_Br5595, pch=20,col=as.factor(colData(spe_sub)$sample_id))



```

## KODAMA analysis

```{r, fig.width=10, fig.height=4}
set.seed(seed)
kk=KODAMA.matrix.parallel(pca_Br5595,
                          spatial = xy,
                          samples=samples,
                          landmarks = 100000,
                          splitting = splitting,
                          ncomp = 50,
                          spatial.resolution = spatial.resolution,
                          n.cores=n.cores,
                          seed = seed)

print("KODAMA finished")
           

config=umap.defaults
config$n_threads = n.cores
config$n_sgd_threads = "auto"

kk_UMAP=KODAMA.visualization(kk,method="UMAP",config=config)
plot(kk_UMAP,pch=20,col=cols_cluster[labels])
     
```
 
```{r, fig.width=7, fig.height=3, warning=FALSE, message=FALSE, echo=FALSE}
svg("output/KODAMA_DLPFC_Br5595.svg")
plot(kk_UMAP,pch=20,col=cols_cluster[labels])
dev.off()
```

## Graph-based clustering

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}

    # Graph-based clustering

g <- bluster::makeSNNGraph(as.matrix(kk_UMAP), k = 20)
g_walk <- igraph::cluster_walktrap(g)
clu <- as.character(igraph::cut_at(g_walk, no = 2))
plot(kk_UMAP,pch=20,col=cols_cluster[as.factor(clu)])



  plot_slide(xy,as.factor(samples),clu,col=cols_cluster,size.dot = 1)

``` 
 
  
```{r, fig.width=14, fig.height=3, warning=FALSE, message=FALSE, echo=FALSE}
svg("output/KODAMA_DLPFC_Br5595_slide.svg",height =  3)
plot_slide(xy,samples,labels,col=cols_cluster,size.dot = 1)
dev.off()
``` 

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}

FB=names(which.min(table(clu)))
selFB=clu!=FB

 # kk_UMAP=kk_UMAP[selFB,]
#  labels=labels[selFB]
#  samples=samples[selFB]
#  xy=xy[selFB,]
    
      
g <- bluster::makeSNNGraph(as.matrix(kk_UMAP[selFB,]), k = graph)
g_walk <- igraph::cluster_walktrap(g)
clu <- as.character(igraph::cut_at(g_walk, no = nclusters))
plot(kk_UMAP[selFB,],pch=20,col=as.factor(clu))




ref=refine_SVM(xy[selFB,],clu,samples[selFB],cost=100)

u=unique(samples[selFB])
for(j in u){
  sel=samples[selFB]==j
  print(mclust::adjustedRandIndex(labels[selFB][sel],ref[sel]))
}
        

      


plot_slide(xy,samples,labels,col=cols_cluster,size.dot = 1)
  

plot_slide(xy[selFB,],samples[selFB],ref,col=cols_cluster,size.dot = 1)
  

kk_UMAP_Br5595=kk_UMAP
samples_Br5595=samples
xy_Br5595=xy
labels_Br5595=labels
ref_Br5595=ref
clu_Br5595=clu

save(top,kk_UMAP_Br5595,samples_Br5595,xy_Br5595,labels_Br5595,subject_names_Br5595,ref_Br5595,clu_Br5595,selFB,file="output/DLFPC-Br5595.RData")

save(top,data_Br5595,pca_Br5595,samples_Br5595,xy_Br5595,labels_Br5595,subject_names_Br5595,selFB,file="data/DLFPC-Br5595-input.RData")



```




# Patient Br5292


```{r, fig.width=10, fig.height=4}

subject_names="Br5292"
nclusters=7

spe_sub <- spe[, colData(spe)$subject ==  subject_names]
dim(spe_sub)

spe_sub <- spe[, colData(spe)$subject ==  subject_names]
sel= subjects ==  subject_names
data_Br5292=data[sel,top[1:gene_number]]
RNA.scaled=scale(data_Br5292)
pca_results <- irlba(A = RNA.scaled, nv = 50)
pca_Br5292 <- pca_results$u %*% diag(pca_results$d)[,1:50]
rownames(pca_Br5292)=rownames(data_Br5292)
colnames(pca_Br5292)=paste("PC",1:50,sep="")
labels=as.factor(colData(spe_sub)$layer_guess_reordered)
names(labels)=rownames(colData(spe_sub))
xy=as.matrix(spatialCoords(spe_sub))
rownames(xy)=rownames(colData(spe_sub))
samples=colData(spe_sub)$sample_id
subject_names_Br5292=colData(spe_sub)$subject
plot(pca_Br5292, pch=20,col=as.factor(colData(spe_sub)$sample_id))



  
```

## KODAMA analysis

```{r, fig.width=10, fig.height=4}
set.seed(seed)
kk=KODAMA.matrix.parallel(pca_Br5292,
                          
                          spatial = xy,
                          samples=samples,
                          landmarks = 100000,
                          splitting = splitting,
                          ncomp = 50,
                          spatial.resolution = spatial.resolution,
                          n.cores=n.cores,
                          seed = seed)
  print("KODAMA finished")
            
     config=umap.defaults
     config$n_threads = n.cores
     config$n_sgd_threads = "auto"
     kk_UMAP=KODAMA.visualization(kk,method="UMAP",config=config)

     plot(kk_UMAP,pch=20,col=as.factor(labels))
     
     
 
```
 


## Graph-based clustering

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}

    # Graph-based clustering

  
        
        g <- bluster::makeSNNGraph(as.matrix(kk_UMAP), k = graph)
        
        g_walk <- igraph::cluster_walktrap(g)
        clu <- as.character(igraph::cut_at(g_walk, no = nclusters))
        
    plot(kk_UMAP,pch=20,col=as.factor(clu))
        
        ref=refine_SVM(xy,clu,samples,cost=100)
        
        u=unique(samples)
        for(j in u){
          sel=samples==j
          
            print(mclust::adjustedRandIndex(labels[sel],ref[sel]))
        }


          g <- bluster::makeSNNGraph(as.matrix(kk_UMAP), k = graph)
          
          g_walk <- igraph::cluster_walktrap(g)
          clu <- as.character(igraph::cut_at(g_walk, no = nclusters))
          ref=refine_SVM(xy,clu,samples,cost=100)
          
          u=unique(samples)
          for(j in u){
            sel=samples==j
            print(mclust::adjustedRandIndex(labels[sel],ref[sel]))
          }
        

      


plot_slide(xy,samples,labels,col=cols_cluster,size.dot = 1)
  
    plot_slide(xy,samples,ref,col=cols_cluster,size.dot = 1)
  

kk_UMAP_Br5292=kk_UMAP
samples_Br5292=samples
xy_Br5292=xy
labels_Br5292=labels
ref_Br5292=ref
clu_Br5292=clu
save(top,kk_UMAP_Br5292,pca_Br5292,samples_Br5292,xy_Br5292,subject_names_Br5292,labels_Br5292,ref_Br5292,clu_Br5292,file="output/DLFPC-Br5292.RData")

save(top,data_Br5292,pca_Br5292,samples_Br5292,xy_Br5292,labels_Br5292,subject_names_Br5292,file="data/DLFPC-Br5292-input.RData")

```









# Patient Br8100


```{r, fig.width=10, fig.height=4}

subject_names="Br8100"

nclusters=7

  spe_sub <- spe[, colData(spe)$subject ==  subject_names]
  dim(spe_sub)
#  spe_sub <- runPCA(spe_sub, 50,subset_row = top[1:gene_number], scale=TRUE)

  #pca=reducedDim(spe_sub,type = "PCA")[,1:50]
  
  
  
    
        
        spe_sub <- spe[, colData(spe)$subject ==  subject_names]
        sel= subjects ==  subject_names
        

        data_Br8100=data[sel,top[1:gene_number]]
        
        RNA.scaled=scale(data_Br8100)
        pca_results <- irlba(A = RNA.scaled, nv = 50)
        pca_Br8100 <- pca_results$u %*% diag(pca_results$d)[,1:50]
        rownames(pca_Br8100)=rownames(data_Br8100)
        colnames(pca_Br8100)=paste("PC",1:50,sep="")
        labels=as.factor(colData(spe_sub)$layer_guess_reordered)
        names(labels)=rownames(colData(spe_sub))
        xy=as.matrix(spatialCoords(spe_sub))
        rownames(xy)=rownames(colData(spe_sub))
        samples=colData(spe_sub)$sample_id
        subject_names_Br8100=colData(spe_sub)$subject

  plot(pca_Br8100, pch=20,col=as.factor(colData(spe_sub)$sample_id))


```

## KODAMA analysis

```{r, fig.width=10, fig.height=4}
set.seed(seed)
kk=KODAMA.matrix.parallel(pca_Br8100,
                          spatial = xy,
                          samples=samples,
                          landmarks = 100000,
                          splitting = splitting,
                          ncomp = 50,
                          spatial.resolution = spatial.resolution,
                          n.cores=n.cores,
                          seed = seed)

  print("KODAMA finished")
            
     config=umap.defaults
     config$n_threads = n.cores
     config$n_sgd_threads = "auto"
     kk_UMAP=KODAMA.visualization(kk,method="UMAP",config=config)

     plot(kk_UMAP,pch=20,col=as.factor(labels))
     
     
```
 


## Graph-based clustering

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}

    # Graph-based clustering

  
        
        g <- bluster::makeSNNGraph(as.matrix(kk_UMAP), k = graph)
        
        g_walk <- igraph::cluster_walktrap(g)
        clu <- as.character(igraph::cut_at(g_walk, no = nclusters))
        ref=refine_SVM(xy,clu,samples,cost=100)
        
        u=unique(samples)
        for(j in u){
          sel=samples==j
          
            print(mclust::adjustedRandIndex(labels[sel],ref[sel]))
        }


          g <- bluster::makeSNNGraph(as.matrix(kk_UMAP), k = graph)
          
          g_walk <- igraph::cluster_walktrap(g)
          clu <- as.character(igraph::cut_at(g_walk, no = nclusters))
          
           plot(kk_UMAP,pch=20,col=as.factor(clu))
          
          ref=refine_SVM(xy,clu,samples,cost=100)
          
          u=unique(samples)
          for(j in u){
            sel=samples==j
            print(mclust::adjustedRandIndex(labels[sel],ref[sel]))
          }
        

      
        


          
 plot_slide(xy,samples,labels,col=cols_cluster,size.dot = 1)
  

    plot_slide(xy,samples,ref,col=cols_cluster,size.dot = 1)
  
kk_UMAP_Br8100=kk_UMAP
samples_Br8100=samples
xy_Br8100=xy
labels_Br8100=labels

ref_Br8100=ref
clu_Br8100=clu 
save(top,kk_UMAP_Br8100,pca_Br8100,samples_Br8100,xy_Br8100,subject_names_Br8100,labels_Br8100,ref_Br8100,clu_Br8100,file="output/DLFPC-Br8100.RData")


save(top,data_Br8100,pca_Br8100,samples_Br8100,xy_Br8100,labels_Br8100,subject_names_Br8100,file="data/DLFPC-Br8100-input.RData")

```
# Saving the results


```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}

result_KODAMA <- list(clusters=list(),
                      tissue_segments=list(),
                      feature_extraction=list(),
                      xy=list(),
                      subjects=list(),
                      samples=list())

result_KODAMA$clusters=c(result_KODAMA$clusters,tapply(ref_Br5292,samples_Br5292,function(x) x))
result_KODAMA$clusters=c(result_KODAMA$clusters,tapply(ref_Br5595,samples_Br5595[selFB],function(x) x))
result_KODAMA$clusters=c(result_KODAMA$clusters,tapply(ref_Br8100,samples_Br8100,function(x) x))

result_KODAMA$tissue_segments=c(result_KODAMA$tissue_segments,tapply(labels_Br5292,samples_Br5292,function(x) x))
result_KODAMA$tissue_segments=c(result_KODAMA$tissue_segments,tapply(labels_Br5595[selFB],samples_Br5595[selFB],function(x) x))
result_KODAMA$tissue_segments=c(result_KODAMA$tissue_segments,tapply(labels_Br8100,samples_Br8100,function(x) x))

result_KODAMA$feature_extraction=c(result_KODAMA$feature_extraction,by(kk_UMAP_Br5292,samples_Br5292,function(x) x))
result_KODAMA$feature_extraction=c(result_KODAMA$feature_extraction,by(kk_UMAP_Br5595[selFB,],samples_Br5595[selFB],function(x) x))
result_KODAMA$feature_extraction=c(result_KODAMA$feature_extraction,by(kk_UMAP_Br8100,samples_Br8100,function(x) x))

result_KODAMA$xy=c(result_KODAMA$xy,by(xy_Br5292,samples_Br5292,function(x) x))
result_KODAMA$xy=c(result_KODAMA$xy,by(xy_Br5595[selFB,],samples_Br5595[selFB],function(x) x))
result_KODAMA$xy=c(result_KODAMA$xy,by(xy_Br8100,samples_Br8100,function(x) x))

result_KODAMA$samples=c(result_KODAMA$samples,tapply(samples_Br5292,samples_Br5292,function(x) x))
result_KODAMA$samples=c(result_KODAMA$samples,tapply(samples_Br5595[selFB],samples_Br5595[selFB],function(x) x))
result_KODAMA$samples=c(result_KODAMA$samples,tapply(samples_Br8100,samples_Br8100,function(x) x))

result_KODAMA$subjects=c(result_KODAMA$subjects,tapply(subject_names_Br5292,samples_Br5292,function(x) x))
result_KODAMA$subjects=c(result_KODAMA$subjects,tapply(subject_names_Br5595[selFB],samples_Br5595[selFB],function(x) x))
result_KODAMA$subjects=c(result_KODAMA$subjects,tapply(subject_names_Br8100,samples_Br8100,function(x) x))


ground_true <- list(clusters=list(),
                    tissue_segments=list(),
                    xy=list(),
                    subjects=list(),
                    samples=list())

ground_true$tissue_segments=c(ground_true$tissue_segments,tapply(labels_Br5292,samples_Br5292,function(x) x))
ground_true$tissue_segments=c(ground_true$tissue_segments,tapply(labels_Br5595,samples_Br5595,function(x) x))
ground_true$tissue_segments=c(ground_true$tissue_segments,tapply(labels_Br8100,samples_Br8100,function(x) x))

ground_true$xy=c(ground_true$xy,by(xy_Br5292,samples_Br5292,function(x) x))
ground_true$xy=c(ground_true$xy,by(xy_Br5595,samples_Br5595,function(x) x))
ground_true$xy=c(ground_true$xy,by(xy_Br8100,samples_Br8100,function(x) x))

ground_true$clusters=ground_true$tissue_segments



ground_true$samples=c(ground_true$samples,tapply(samples_Br5292,samples_Br5292,function(x) x))
ground_true$samples=c(ground_true$samples,tapply(samples_Br5595,samples_Br5595,function(x) x))
ground_true$samples=c(ground_true$samples,tapply(samples_Br8100,samples_Br8100,function(x) x))

ground_true$subjects=c(ground_true$subjects,tapply(subject_names_Br5292,samples_Br5292,function(x) x))
ground_true$subjects=c(ground_true$subjects,tapply(subject_names_Br5595,samples_Br5595,function(x) x))
ground_true$subjects=c(ground_true$subjects,tapply(subject_names_Br8100,samples_Br8100,function(x) x))


save(result_KODAMA,ground_true,file="output/KODAMA-results.RData")



source("code/DLFPC_preprocessing.R")
```




# 12 Slides


PCA and HARMONY

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}


 spe <- runPCA(spe, 50,subset_row = top[1:gene_number], scale=TRUE)


 subjects=colData(spe)$subject
 labels=as.factor(colData(spe)$layer_guess_reordered)
 xy=as.matrix(spatialCoords(spe))
 samples=colData(spe)$sample_id
 
  
 

 spe <- RunHarmony(spe, "subject",lambda=NULL)
 pca=reducedDim(spe,type = "HARMONY")[,1:50]
 
 plot(pca, pch=20,col=as.factor(colData(spe_sub)$sample_id))
```


KODAMA

```{r, fig.width=10, fig.height=4}
set.seed(seed)
kk=KODAMA.matrix.parallel(pca,
                          spatial = xy,
                          samples=samples,
                          landmarks = 100000,
                          splitting = splitting,
                          ncomp = 50,
                          spatial.resolution = spatial.resolution,
                          n.cores=n.cores,
                          seed = seed)
print("KODAMA finished")
           

config=umap.defaults
config$n_threads = n.cores
config$n_sgd_threads = "auto"

kk_UMAP=KODAMA.visualization(kk,method="UMAP",config=config)
plot(kk_UMAP,pch=20,col=as.factor(labels))

df <- data.frame(kk_UMAP[,1:2], tissue=labels,check.names = FALSE)
 
plot1 = ggplot(df, aes(`Dimension 1`, `Dimension 2`, color = tissue)) +labs(title="KODAMA") +
     geom_point(size = 1) +
     theme_bw() + theme(legend.position = "bottom",
                        legend.text = element_text(size = 20),
                        legend.title = element_text(size = 20),
                        axis.title = element_text(size = 22),       # x and y axis labels
                        axis.text = element_text(size = 16),        # tick labels
                        plot.title = element_text(size = 26, face = "bold", hjust = 0) )+
     scale_color_manual("Domain", values = cols_cluster) +
     
     guides(color = guide_legend(nrow = 1,override.aes = list(size = 10)))
plot1

```


```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
png("output/Figures/DLPFC/Fig S9.png",width = 1200,height = 1200)
print(plot1)
dev.off()

save(kk_UMAP,samples,xy,labels,file="output/DLFPC-All.RData")
```


CLUSTER

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
  g <- bluster::makeSNNGraph(as.matrix(kk_UMAP), k = graph)
  g_walk <- igraph::cluster_walktrap(g)
  clu <- as.character(igraph::cut_at(g_walk, no = 7))
  plot(kk_UMAP,pch=20,col=cols_cluster[as.factor(clu)]) 

  plot_slide(xy,samples,clu,col=cols_cluster,size.dot = 1)

  mito=colData(spe)$subsets_mito_percent
  sel_local=(labels=="Layer3" | labels=="Layer4") & 
    (samples=="151669" | samples=="151670" | samples=="151671" | samples=="151672") &
    (clu==1 | clu==6)

  
   
   
library(ggpubr)
library(ggplot2)

df=data.frame(mito=mito[sel_local],labels=as.character(clu[sel_local]))

my_comparisons=list(c("1","6"))
Nplot1=ggboxplot(df, x = "labels", y = "mito", width = 0.8,palette = cols_cluster[c(1,6)] ,las=2,
                 fill="labels",
                 shape=21)+  
    ylab("Mitochondrial gene percentage")+ 
    xlab("")+
    
    stat_compare_means(comparisons = my_comparisons,method="wilcox.test")+
    theme(legend.position = "none",plot.margin = unit(c(2,1,1,1), "cm"))

Nplot1

   
```


```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
pdf("output/Figures/DLPFC/Fig S10.pdf")
print(Nplot1)
dev.off()
png("output/Figures/DLPFC/Fig S10.png")
print(Nplot1)
dev.off()
save(kk_UMAP,data,samples,xy,labels,clu,mito,file="output/DLFPC-All.RData")
```


CLUSTER

We removed the spots with uncertain quality level.

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
    
  sel_rem=which((clu %in% names(sort(table(clu)))[1:2]) |
                 rownames(kk_UMAP) %in% rownames(xy_Br5595[!selFB,]))

  kk_UMAP_clear=kk_UMAP[-sel_rem,]
  labels_clear=labels[-sel_rem]
  samples_clear=samples[-sel_rem]
  xy_clear=xy[-sel_rem,]
  data_clear=data[-sel_rem,]
  subjects_clear=subjects[-sel_rem]
  
  clu_clear=kmeans(kk_UMAP_clear,7,nstart = 100)$cluster
  plot(kk_UMAP_clear,col=cols_cluster[labels_clear],pch=20)
```  
  
```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
  svg("output/KODAMA_DLPFC_All_original.svg")
  plot(kk_UMAP_clear,col=cols_cluster[labels_clear],pch=20)
dev.off()
```

KODAMA plot colored by cluster.

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
   
  plot(kk_UMAP_clear,col=cols_cluster[clu_clear],pch=20)


u=unique(samples)
for(i in 1:length(u)){
  sel=samples==u[i]
  print(adjustedRandIndex(labels_clear[sel],clu_clear[sel]))
}

ref=refine_SVM(xy_clear,clu_clear,samples_clear,cost=100)
names(ref)=rownames(data_clear)
names(labels_clear)=rownames(data_clear)

u=unique(samples)
for(i in 1:length(u)){
  sel=samples[-sel_rem]==u[i]
  print(adjustedRandIndex(labels_clear[sel],ref[sel]))
}

```
  
  
```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
save(kk_UMAP,data,samples,xy,labels,clu,mito,
     labels_clear,ref,subjects_clear,samples_clear,
     kk_UMAP_clear,data_clear,xy_clear,ref,clu_clear,
     file="output/DLFPC-All.RData")



```


# Trajectory color


```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}


d <- slingshot(kk_UMAP_clear, clusterLabels = clu_clear)
trajectory=d@metadata$curves$Lineage1$s
k=Rnanoflann::nn(trajectory,kk_UMAP_clear,1)
map_color=rainbow(nrow(trajectory))[k$indices]
plot(kk_UMAP_clear,pch=20,col=map_color)
points(trajectory,pch=21,bg="white",col=2,lwd=2)
plot_slide(xy_clear,samples_clear,k$indices,col=rainbow(nrow(trajectory)),size.dot = 1)
```


```{r, fig.width=14, fig.height=3, warning=FALSE, message=FALSE, echo=FALSE}
png("output/Figures/DLPFC/Fig 3h.png",width = 4000,height = 3300,bg = "transparent")
plot_slide_minimal(xy_clear,samples_clear,k$indices,col=rainbow(nrow(trajectory)),size.dot = 5,nrow=3)
dev.off()
```


The cluster number is reorder based on the trajectory

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
oo=order(tapply(k$indices,ref,mean))
tra=1:7
names(tra)=oo
ref_ordered=tra[as.character(ref)]
   
plot_slide(xy_clear,samples_clear,ref_ordered,col=cols_cluster,size.dot = 1)

```


```{r, fig.width=14, fig.height=3, warning=FALSE, message=FALSE, echo=FALSE}
png("output/Figures/DLPFC/Fig 3g.png",width = 12000,height = 1100,bg = "transparent")
plot_slide_minimal(xy_clear,samples_clear,ref_ordered,col=cols_cluster,size.dot = 5)
dev.off()
```


```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
png("output/Figures/DLPFC/RGB_DLPFC.png",width = 2000,height = 2000)
plot(kk_UMAP_clear,pch=20,col=map_color,axes=FALSE,xlab="",ylab="")
points(trajectory,pch=21,bg="white",col=2,lwd=3,cex=3)
dev.off()

svg("output/DLPFC_all_cluster.svg",height = 3)
plot_slide(xy_clear,samples_clear,ref_ordered,col=cols_cluster,size.dot = 1)
dev.off()
```


The variable to be used into the deep learning approach

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}

save(ref_ordered,samples_clear,subjects_clear,labels_clear,
     file="output/DLFPC-variablesXdeeplearning.RData")
```





```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE,results = "hide"}


library(ggalluvial)


plot_sankey_subject <- function(subject_id, labels_clear, ref_ordered, subjects_clear, cols_cluster) {
  sel_sub <- subjects_clear == subject_id
  al <- data.frame(
    expand.grid(list(levels(labels_clear), 1:7)),
    freq = as.numeric(table(labels_clear[sel_sub], ref_ordered[sel_sub]))
  )
  al$Var2 <- as.factor(al$Var2)
  
  ggplot(data = al, aes(axis1 = Var1, axis2 = Var2, y = freq)) +
    geom_alluvium(aes(fill = Var2)) +
    geom_stratum() +
    scale_fill_manual(values = cols_cluster) +
    geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
    theme_void()
}

Sankey_Br5595 <- plot_sankey_subject("Br5595", labels_clear, ref_ordered, subjects_clear, cols_cluster)
Sankey_Br5595
Sankey_Br5292 <- plot_sankey_subject("Br5292", labels_clear, ref_ordered, subjects_clear, cols_cluster)
Sankey_Br5292
Sankey_Br8100 <- plot_sankey_subject("Br8100", labels_clear, ref_ordered, subjects_clear, cols_cluster)
Sankey_Br8100


```


```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
library(patchwork)

# Remove individual legends
Sankey_Br5595 <- Sankey_Br5595 + theme(legend.position = "none")
Sankey_Br5292 <- Sankey_Br5292 + theme(legend.position = "none")
# Keep legend in the last plot (or any one of them)
Sankey_Br8100 <- Sankey_Br8100 + theme(legend.position = "right")

# Combine horizontally and collect the legend
combined_plot <- (Sankey_Br5292 | Sankey_Br5595 | Sankey_Br8100) + plot_layout(guides = "collect")

pdf("output/Figures/DLPFC/Fig 3i.pdf",width = 8)
print(combined_plot)
dev.off()


```




```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE,results = "hide"}
subclusters=sort(names(sort(table(ref_ordered,labels_clear)[,"Layer3"],decreasing = TRUE)[1:3]))
print(subclusters)




sa_sel=(ref_ordered %in% subclusters) & (samples_clear %in% c("151669","151670","151671","151672")) & c(labels_clear %in% "Layer3")


plot_slide(xy_clear[sa_sel,],samples_clear[sa_sel],ref_ordered[sa_sel],col=cols_cluster,size.dot = 1)

s1=rownames(data_clear)[which(ref_ordered==subclusters[1] & sa_sel)]
s2=rownames(data_clear)[which(ref_ordered==subclusters[2] & sa_sel)]
s3=rownames(data_clear)[which(ref_ordered==subclusters[3] & sa_sel)]



data_clear_N=data_clear[c(s1,s2,s3),]
data_clear_N=10*t(t(data_clear_N)/colMaxs(data_clear_N))
colnames(data_clear_N)=genes[colnames(data_clear_N)]

lab_S123 <- factor(rep(c("Group1", "Group2", "Group3"), times = c(length(s1), length(s2), length(s3))),levels=c("Group1","Group2","Group3"))
da123=multi_analysis(data_clear_N[c(s1,s2,s3),],lab_S123,FUN="continuous.test",range="95%CI")


lab_S12 <- factor(rep(c("Group1", "Group2"), times = c(length(s1), length(s2))))
lab_S13 <- factor(rep(c("Group1", "Group3"), times = c(length(s1), length(s3))))
da12=multi_analysis(data_clear_N[c(s1,s2),],lab_S12,FUN="continuous.test",alternative = "greater")
da13=multi_analysis(data_clear_N[c(s1,s3),],lab_S13,FUN="continuous.test",alternative = "greater")

rank=order(-log(1+as.numeric(da12$`p-value`))-log(1+as.numeric(da13$`p-value`)),decreasing = TRUE)[1:50]

df=data.frame(da123[,1],
           sublayer1=da123[,2],
           sublayer3=da123[,3],
           sublayer5=da123[,4],
           L1vsL3_pvalue=da12[,4],
           L1vsL3_FDR=da12[,5],
           L1vsL5_pvalue=da13[,4],
           L1vsL5_FDR=da13[,5])[rank,]


write.csv(df,"output/subclusters1.csv")
####

lab_S21 <- factor(rep(c("Group2", "Group1"), times = c(length(s2), length(s1))),levels=c("Group2","Group1"))
lab_S23 <- factor(rep(c("Group2", "Group3"), times = c(length(s2), length(s3))),levels=c("Group2","Group3"))
da21=multi_analysis(data_clear_N[c(s2,s1),],lab_S21,FUN="continuous.test",alternative = "greater",range="95%CI")
da23=multi_analysis(data_clear_N[c(s2,s3),],lab_S23,FUN="continuous.test",alternative = "greater",range="95%CI")

rank=order(-log(1+as.numeric(da21$`p-value`))-log(1+as.numeric(da23$`p-value`)),decreasing = TRUE)[1:50]

df=data.frame(da123[,1],
           sublayer1=da123[,2],
           sublayer3=da123[,3],
           sublayer5=da123[,4],
           L3vsL1_pvalue=da21[,4],
           L3vsL1_FDR=da21[,5],
           L3vsL5_pvalue=da23[,4],
           L3vsL5_FDR=da23[,5])[rank,]


write.csv(df,"output/subclusters2.csv")

####


lab_S31 <- factor(rep(c("Group3", "Group1"), times = c(length(s3), length(s1))),levels=c("Group3","Group1"))
lab_S32 <- factor(rep(c("Group3", "Group2"), times = c(length(s3), length(s2))),levels=c("Group3","Group2"))
da31=multi_analysis(data_clear_N[c(s3,s1),],lab_S31,FUN="continuous.test",alternative = "greater")
da32=multi_analysis(data_clear_N[c(s3,s2),],lab_S32,FUN="continuous.test",alternative = "greater")

rank=order(-log(1+as.numeric(da31$`p-value`))-log(1+as.numeric(da32$`p-value`)),decreasing = TRUE)[1:50]

df=data.frame(da123[,1],
           sublayer1=da123[,2],
           sublayer3=da123[,3],
           sublayer5=da123[,4],
           L5vsL1_pvalue=da31[,4],
           L5vsL1_FDR=da31[,5],
           L5vsL3_pvalue=da32[,4],
           L5vsL3_FDR=da32[,5])[rank,]


write.csv(df,"output/subclusters3.csv")


```





```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
#source("code/DLPFC_comparison.R")
```

