---
title: "DLPFC"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

# Introduction
Here, we apply KODAMA to analyze the human dorsolateral prefrontal cortex (DLPFC) data by 10x Visium from [Maynard et al., 2021](https://www.nature.com/articles/s41593-020-00787-0). The links to download the raw data and H&E full resolution images can be found in the [LieberInstitute/spatialLIBD](https://github.com/LieberInstitute/spatialLIBD) github page.

# Loading the required libraries
```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
library("nnSVG")
library("scater")
library("scran")
library("scry")
library("SPARK")
library("harmony")
library("Seurat")
library("spatialLIBD")
library("KODAMAextra")
library("mclust")
library("slingshot")
library("irlba")
library("Rnanoflann")
library("ggpubr")
```

# Download the dataset
```{r, eval=FALSE}
spe <- fetch_data(type = 'spe')
```

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
load("../DLFPC.RData")
```

## Extract the metadata information
```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
n.cores=40
splitting = 100
spatial.resolution = 0.3
aa_noise=3
gene_number=2000
graph = 20
seed=543210


set.seed(seed)
ID=unlist(lapply(strsplit(rownames(colData(spe)),"-"),function(x) x[1]))
samples=colData(spe)$sample_id
rownames(colData(spe))=paste(ID,samples,sep="-")

txtfile=paste(splitting,spatial.resolution,aa_noise,2,gene_number,sep="_")

sample_names=c("151507",
               "151508",
               "151509",
               "151510",
               "151669",
               "151670",
               "151671",
               "151672",
               "151673",
               "151674",
               "151675",
               "151676")
subject_names= c("Br5292","Br5595", "Br8100")
metaData = SingleCellExperiment::colData(spe)
expr = SingleCellExperiment::counts(spe)
sample_names <- paste0("sample_", unique(colData(spe)$sample_id))
sample_names <-  unique(colData(spe)$sample_id)
dim(spe)
# identify mitochondrial genes
is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
table(is_mito)

# calculate per-spot QC metrics
spe <- addPerCellQC(spe, subsets = list(mito = is_mito))

# select QC thresholds
qc_lib_size <- colData(spe)$sum < 500
qc_detected <- colData(spe)$detected < 250
qc_mito <- colData(spe)$subsets_mito_percent > 30
qc_cell_count <- colData(spe)$cell_count > 12

# spots to discard
discard <- qc_lib_size | qc_detected | qc_mito | qc_cell_count
table(discard)
colData(spe)$discard <- discard
# filter low-quality spots
spe <- spe[, !colData(spe)$discard]
dim(spe)


spe <- filter_genes(
  spe,
  filter_genes_ncounts = 2,   #ncounts
  filter_genes_pcspots = 0.5,
  filter_mito = TRUE
)

dim(spe)

sel= !is.na(colData(spe)$layer_guess_reordered)
spe = spe[,sel]
dim(spe)

spe <- computeLibraryFactors(spe)
spe <- logNormCounts(spe)

 subjects=colData(spe)$subject
 labels=as.factor(colData(spe)$layer_guess_reordered)
 xy=as.matrix(spatialCoords(spe))
 samples=colData(spe)$sample_id
 
 cols_cluster <- c("#0000b6", "#81b29a", "#f2cc8f","#e07a5f",
          "#cc00b6", "#81ccff", "#33b233")

  plot_slide(xy,samples,labels,col=cols_cluster)
  
```

```{r, fig.width=14, fig.height=3, warning=FALSE, message=FALSE, echo=FALSE}
svg("output/DLPFC1.svg",height = 3)
plot_slide(xy,samples,labels,col=cols_cluster)
dev.off()
```

## Gene selection
The identification of genes that display spatial expression patterns is performed using the SPARKX method ([Zhu et al. (2021)](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02404-0)). The genes are ranked based on the median value of the logarithm value of the p-value obtained in each slide individually.


```{r, fig.width=10, fig.height=4,results = "hide"}


top=multi_SPARKX(spe,n.cores=n.cores)

data=as.matrix(t(logcounts(spe)[top[1:gene_number],]))
samples=colData(spe)$sample_id
labels=as.factor(colData(spe)$layer_guess_reordered)
names(labels)=rownames(colData(spe))
subjects=colData(spe)$subject
  
```





# Patient Br5595


```{r, fig.width=10, fig.height=4}

subject_names="Br5595"
nclusters=5

spe_sub <- spe[, colData(spe)$subject ==  subject_names]
 # subjects=colData(spe_sub)$subject
dim(spe_sub)
#  spe_sub <- runPCA(spe_sub, 50,subset_row = top[1:gene_number], scale=TRUE)
#pca=reducedDim(spe_sub,type = "PCA")[,1:50]
  
spe_sub <- spe[, colData(spe)$subject ==  subject_names]
sel= subjects ==  subject_names
        
data_Br5595=data[sel,top[1:gene_number]]
        
RNA.scaled=scale(data_Br5595)
pca_results <- irlba(A = RNA.scaled, nv = 50)
pca_Br5595 <- pca_results$u %*% diag(pca_results$d)[,1:50]
rownames(pca_Br5595)=rownames(data_Br5595)
colnames(pca_Br5595)=paste("PC",1:50,sep="")
labels=as.factor(colData(spe_sub)$layer_guess_reordered)
names(labels)=rownames(colData(spe_sub))
xy=as.matrix(spatialCoords(spe_sub))
rownames(xy)=rownames(colData(spe_sub))
samples=colData(spe_sub)$sample_id


plot(pca_Br5595, pch=20,col=as.factor(colData(spe_sub)$sample_id))

```

## KODAMA analysis

```{r, fig.width=10, fig.height=4}
set.seed(seed)
kk=KODAMA.matrix.parallel(pca_Br5595,
                          spatial = xy,
                          samples=samples,
                          FUN= "fastpls" ,
                          landmarks = 100000,
                          splitting = splitting,
                          ncomp = 50,
                          spatial.resolution = spatial.resolution,
                          n.cores=n.cores,
                          seed = seed)

print("KODAMA finished")
           

config=umap.defaults
config$n_threads = n.cores
config$n_sgd_threads = "auto"

kk_UMAP=KODAMA.visualization(kk,method="UMAP",config=config)
plot(kk_UMAP,pch=20,col=cols_cluster[labels])
     
```
 
```{r, fig.width=7, fig.height=3, warning=FALSE, message=FALSE, echo=FALSE}
svg("output/KODAMA_DLPFC_Br5595.svg")
plot(kk_UMAP,pch=20,col=cols_cluster[labels])
dev.off()
```

## Graph-based clustering

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}

    # Graph-based clustering

g <- bluster::makeSNNGraph(as.matrix(kk_UMAP), k = 20)
g_walk <- igraph::cluster_walktrap(g)
clu <- as.character(igraph::cut_at(g_walk, no = 2))
plot(kk_UMAP,pch=20,col=cols_cluster[as.factor(clu)])



  plot_slide(xy,as.factor(samples),clu,col=cols_cluster)

``` 
 
  
```{r, fig.width=14, fig.height=3, warning=FALSE, message=FALSE, echo=FALSE}
svg("output/KODAMA_DLPFC_Br5595_slide.svg",height =  3)
plot_slide(xy,samples,labels,col=cols_cluster)
dev.off()
``` 

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}

FB=names(which.min(table(clu)))
selFB=clu!=FB

 # kk_UMAP=kk_UMAP[selFB,]
#  labels=labels[selFB]
#  samples=samples[selFB]
#  xy=xy[selFB,]
    
      
g <- bluster::makeSNNGraph(as.matrix(kk_UMAP[selFB,]), k = graph)
g_walk <- igraph::cluster_walktrap(g)
clu <- as.character(igraph::cut_at(g_walk, no = nclusters))
plot(kk_UMAP[selFB,],pch=20,col=as.factor(clu))




ref=refine_SVM(xy[selFB,],clu,samples[selFB],cost=100)

u=unique(samples[selFB])
for(j in u){
  sel=samples[selFB]==j
  print(mclust::adjustedRandIndex(labels[selFB][sel],ref[sel]))
}
        
      ###########
      


plot_slide(xy,samples,labels,col=cols_cluster)
  

plot_slide(xy[selFB,],samples[selFB],ref,col=cols_cluster)
  

kk_UMAP_Br5595=kk_UMAP
samples_Br5595=samples
xy_Br5595=xy
labels_Br5595=labels
subject_names_Br5595=subject_names
ref_Br5595=ref
clu_Br5595=clu

save(kk_UMAP_Br5595,samples_Br5595,xy_Br5595,labels_Br5595,subject_names_Br5595,ref_Br5595,clu_Br5595,selFB,file="output/DLFPC-Br5595.RData")

save(data_Br5595,pca_Br5595,samples_Br5595,xy_Br5595,labels_Br5595,subject_names_Br5595,selFB,file="data/DLFPC-Br5595-input.RData")



```





# Patient Br5292


```{r, fig.width=10, fig.height=4}

subject_names="Br5292"
nclusters=7

  spe_sub <- spe[, colData(spe)$subject ==  subject_names]
  dim(spe_sub)
#  spe_sub <- runPCA(spe_sub, 50,subset_row = top[1:gene_number], scale=TRUE)

  #pca=reducedDim(spe_sub,type = "PCA")[,1:50]
  
  
  
    
        
        spe_sub <- spe[, colData(spe)$subject ==  subject_names]
        sel= subjects ==  subject_names
        

        data_Br5292=data[sel,top[1:gene_number]]
        
        RNA.scaled=scale(data_Br5292)
        pca_results <- irlba(A = RNA.scaled, nv = 50)
        pca_Br5292 <- pca_results$u %*% diag(pca_results$d)[,1:50]
        rownames(pca_Br5292)=rownames(data_Br5292)
        colnames(pca_Br5292)=paste("PC",1:50,sep="")
        labels=as.factor(colData(spe_sub)$layer_guess_reordered)
        names(labels)=rownames(colData(spe_sub))
        xy=as.matrix(spatialCoords(spe_sub))
        rownames(xy)=rownames(colData(spe_sub))
        samples=colData(spe_sub)$sample_id
        
  plot(pca_Br5292, pch=20,col=as.factor(colData(spe_sub)$sample_id))


```

## KODAMA analysis

```{r, fig.width=10, fig.height=4}
set.seed(seed)
kk=KODAMA.matrix.parallel(pca_Br5292,
                          
                          spatial = xy,
                          samples=samples,
                          FUN= "fastpls" ,
                          landmarks = 100000,
                          splitting = splitting,
                          ncomp = 50,
                          spatial.resolution = spatial.resolution,
                          n.cores=n.cores,
                          seed = seed)
  print("KODAMA finished")
            
     config=umap.defaults
     config$n_threads = n.cores
     config$n_sgd_threads = "auto"
     kk_UMAP=KODAMA.visualization(kk,method="UMAP",config=config)

     plot(kk_UMAP,pch=20,col=as.factor(labels))
     
     
 
```
 


## Graph-based clustering

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}

    # Graph-based clustering

  
        
        g <- bluster::makeSNNGraph(as.matrix(kk_UMAP), k = graph)
        
        g_walk <- igraph::cluster_walktrap(g)
        clu <- as.character(igraph::cut_at(g_walk, no = nclusters))
        
    plot(kk_UMAP,pch=20,col=as.factor(clu))
        
        ref=refine_SVM(xy,clu,samples,cost=100)
        
        u=unique(samples)
        for(j in u){
          sel=samples==j
          
            print(mclust::adjustedRandIndex(labels[sel],ref[sel]))
        }
        ###########

          g <- bluster::makeSNNGraph(as.matrix(kk_UMAP), k = graph)
          
          g_walk <- igraph::cluster_walktrap(g)
          clu <- as.character(igraph::cut_at(g_walk, no = nclusters))
          ref=refine_SVM(xy,clu,samples,cost=100)
          
          u=unique(samples)
          for(j in u){
            sel=samples==j
            print(mclust::adjustedRandIndex(labels[sel],ref[sel]))
          }
        
      ###########
      


plot_slide(xy,samples,labels,col=cols_cluster)
  
    plot_slide(xy,samples,ref,col=cols_cluster)
  

kk_UMAP_Br5292=kk_UMAP
samples_Br5292=samples
xy_Br5292=xy
labels_Br5292=labels
subject_names_Br5292=subject_names
ref_Br5292=ref
clu_Br5292=clu
save(kk_UMAP_Br5292,pca_Br5292,samples_Br5292,xy_Br5292,subject_names_Br5292,labels_Br5292,ref_Br5292,clu_Br5292,file="output/DLFPC-Br5292.RData")



save(data_Br5292,pca_Br5292,samples_Br5292,xy_Br5292,labels_Br5292,subject_names_Br5292,file="data/DLFPC-Br5292-input.RData")

```





# Patient Br8100


```{r, fig.width=10, fig.height=4}

subject_names="Br8100"

nclusters=7

  spe_sub <- spe[, colData(spe)$subject ==  subject_names]
  dim(spe_sub)
#  spe_sub <- runPCA(spe_sub, 50,subset_row = top[1:gene_number], scale=TRUE)

  #pca=reducedDim(spe_sub,type = "PCA")[,1:50]
  
  
  
    
        
        spe_sub <- spe[, colData(spe)$subject ==  subject_names]
        sel= subjects ==  subject_names
        

        data_Br8100=data[sel,top[1:gene_number]]
        
        RNA.scaled=scale(data_Br8100)
        pca_results <- irlba(A = RNA.scaled, nv = 50)
        pca_Br8100 <- pca_results$u %*% diag(pca_results$d)[,1:50]
        rownames(pca_Br8100)=rownames(data_Br8100)
        colnames(pca_Br8100)=paste("PC",1:50,sep="")
        labels=as.factor(colData(spe_sub)$layer_guess_reordered)
        names(labels)=rownames(colData(spe_sub))
        xy=as.matrix(spatialCoords(spe_sub))
        rownames(xy)=rownames(colData(spe_sub))
        samples=colData(spe_sub)$sample_id
        
  plot(pca_Br8100, pch=20,col=as.factor(colData(spe_sub)$sample_id))


```

## KODAMA analysis

```{r, fig.width=10, fig.height=4}
set.seed(seed)
kk=KODAMA.matrix.parallel(pca_Br8100,
                          spatial = xy,
                          samples=samples,
                          FUN= "fastpls" ,
                          landmarks = 100000,
                          splitting = splitting,
                          ncomp = 50,
                          spatial.resolution = spatial.resolution,
                          n.cores=n.cores,
                          seed = seed)

  print("KODAMA finished")
            
     config=umap.defaults
     config$n_threads = n.cores
     config$n_sgd_threads = "auto"
     kk_UMAP=KODAMA.visualization(kk,method="UMAP",config=config)

     plot(kk_UMAP,pch=20,col=as.factor(labels))
     
     
```
 


## Graph-based clustering

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}

    # Graph-based clustering

  
        
        g <- bluster::makeSNNGraph(as.matrix(kk_UMAP), k = graph)
        
        g_walk <- igraph::cluster_walktrap(g)
        clu <- as.character(igraph::cut_at(g_walk, no = nclusters))
        ref=refine_SVM(xy,clu,samples,cost=100)
        
        u=unique(samples)
        for(j in u){
          sel=samples==j
          
            print(mclust::adjustedRandIndex(labels[sel],ref[sel]))
        }
        ###########

          g <- bluster::makeSNNGraph(as.matrix(kk_UMAP), k = graph)
          
          g_walk <- igraph::cluster_walktrap(g)
          clu <- as.character(igraph::cut_at(g_walk, no = nclusters))
          
           plot(kk_UMAP,pch=20,col=as.factor(clu))
          
          ref=refine_SVM(xy,clu,samples,cost=100)
          
          u=unique(samples)
          for(j in u){
            sel=samples==j
            print(mclust::adjustedRandIndex(labels[sel],ref[sel]))
          }
        
      ###########
      
        


          
 plot_slide(xy,samples,labels,col=cols_cluster)
  

    plot_slide(xy,samples,ref,col=cols_cluster)
  
kk_UMAP_Br8100=kk_UMAP
samples_Br8100=samples
xy_Br8100=xy
labels_Br8100=labels
subject_names_Br8100=subject_names
ref_Br8100=ref
clu_Br8100=clu 
save(kk_UMAP_Br8100,pca_Br8100,samples_Br8100,xy_Br8100,subject_names_Br8100,labels_Br8100,ref_Br8100,clu_Br8100,file="output/DLFPC-Br8100.RData")


save(data_Br8100,pca_Br8100,samples_Br8100,xy_Br8100,labels_Br8100,subject_names_Br8100,file="data/DLFPC-Br8100-input.RData")

```
# Saving the results


```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}

results_KODAMA <- list()
results_KODAMA$clusters <- list()
results_KODAMA$labels <- list()
results_KODAMA$feature_extraction <- list()
results_KODAMA$xy <- list()


results_KODAMA$clusters=c(results_KODAMA$clusters,tapply(ref_Br5292,samples_Br5292,function(x) x))
results_KODAMA$clusters=c(results_KODAMA$clusters,tapply(ref_Br5595,samples_Br5595[selFB],function(x) x))
results_KODAMA$clusters=c(results_KODAMA$clusters,tapply(ref_Br8100,samples_Br8100,function(x) x))

results_KODAMA$labels=c(results_KODAMA$labels,tapply(labels_Br5292,samples_Br5292,function(x) x))
results_KODAMA$labels=c(results_KODAMA$labels,tapply(labels_Br5595[selFB],samples_Br5595[selFB],function(x) x))
results_KODAMA$labels=c(results_KODAMA$labels,tapply(labels_Br8100,samples_Br8100,function(x) x))

results_KODAMA$feature_extraction=c(results_KODAMA$feature_extraction,by(kk_UMAP_Br5292,samples_Br5292,function(x) x))
results_KODAMA$feature_extraction=c(results_KODAMA$feature_extraction,by(kk_UMAP_Br5595[selFB,],samples_Br5595[selFB],function(x) x))
results_KODAMA$feature_extraction=c(results_KODAMA$feature_extraction,by(kk_UMAP_Br8100,samples_Br8100,function(x) x))

results_KODAMA$xy=c(results_KODAMA$xy,by(xy_Br5292,samples_Br5292,function(x) x))
results_KODAMA$xy=c(results_KODAMA$xy,by(xy_Br5595[selFB,],samples_Br5595[selFB],function(x) x))
results_KODAMA$xy=c(results_KODAMA$xy,by(xy_Br8100,samples_Br8100,function(x) x))

save(results_KODAMA,file="output/KODAMA-results.RData")

```


#seurat list preprocessing

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
source("code/DLFPC_preprocessing.R")
```




# 12 Slides


PCA and HARMONY

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}


 dim(spe_sub)
 spe <- runPCA(spe, 50,subset_row = top[1:gene_number], scale=TRUE)


 subjects=colData(spe)$subject
 labels=as.factor(colData(spe)$layer_guess_reordered)
 xy=as.matrix(spatialCoords(spe))
 samples=colData(spe)$sample_id
 
  
 

 spe <- RunHarmony(spe, "subject",lambda=NULL)
 pca=reducedDim(spe,type = "HARMONY")[,1:50]
 
 plot(pca, pch=20,col=as.factor(colData(spe_sub)$sample_id))
```


KODAMA

```{r, fig.width=10, fig.height=4}
set.seed(seed)
kk=KODAMA.matrix.parallel(pca,
                          spatial = xy,
                          samples=samples,
                          FUN= "fastpls" ,
                          landmarks = 100000,
                          splitting = splitting,
                          ncomp = 50,
                          spatial.resolution = spatial.resolution,
                          n.cores=n.cores,
                          seed = seed)
print("KODAMA finished")
           

config=umap.defaults
config$n_threads = n.cores
config$n_sgd_threads = "auto"

kk_UMAP=KODAMA.visualization(kk,method="UMAP",config=config)
plot(kk_UMAP,pch=20,col=as.factor(labels))
```


```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
save(kk_UMAP,samples,xy,labels,file="output/DLFPC-All.RData")
```


CLUSTER

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
        g <- bluster::makeSNNGraph(as.matrix(kk_UMAP), k = graph)
        g_walk <- igraph::cluster_walktrap(g)
        clu <- as.character(igraph::cut_at(g_walk, no = 7))
        plot(kk_UMAP,pch=20,col=as.factor(clu)) 


    plot_slide(xy,samples,clu,col=cols_cluster)

    mito=colData(spe)$subsets_mito_percent
    sel_local=(labels=="Layer3" | labels=="Layer4") & 
      (samples=="151669" | samples=="151670" | samples=="151671" | samples=="151672") &
      (clu==1 | clu==6)
  boxplot(mito[sel_local]~clu[sel_local],col=cols_cluster[c(1,6)] ,ylab="mito percent")
   wilcox.test(mito[sel_local]~clu[sel_local])
```


```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
save(kk_UMAP,samples,xy,labels,clu,mito,file="output/DLFPC-All.RData")
```


CLUSTER



```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
    
  sel_rem=which(clu %in% names(sort(table(clu)))[1:2])

  kk_UMAP_clear=kk_UMAP[-sel_rem,]
  labels_clear=labels[-sel_rem]
  samples_clear=samples[-sel_rem]
  xy_clear=xy[-sel_rem,]
  data_clear=data[-sel_rem,]
  subjects_clear=subjects[-sel_rem]
  
  clu=kmeans(kk_UMAP_clear,7,nstart = 100)$cluster
  plot(kk_UMAP_clear,col=cols_cluster[labels_clear],pch=20)
```  
  
```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
  svg("output/KODAMA_DLPFC_All_original.svg")
  plot(kk_UMAP_clear,col=cols_cluster[labels_clear],pch=20)
dev.off()
```

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
   
  plot(kk_UMAP_clear,col=cols_cluster[clu],pch=20)


u=unique(samples)
for(i in 1:length(u)){
  sel=samples==u[i]
  print(adjustedRandIndex(labels_clear[sel],clu[sel]))
}

ref=refine_SVM(xy_clear,clu,samples_clear,cost=100)
names(ref)=rownames(data_clear)
names(labels_clear)=rownames(data_clear)
save(labels_clear,ref,subjects_clear,samples_clear,file="output/DL.RData")

u=unique(samples)
for(i in 1:length(u)){
  sel=samples[-sel_rem]==u[i]
  print(adjustedRandIndex(labels_clear[sel],ref[sel]))
}
plot_slide(xy_clear,samples_clear,ref,col=cols_cluster)

```
  
  

```{r, fig.width=14, fig.height=3, warning=FALSE, message=FALSE, echo=FALSE}
svg("output/DLPFC_all_cluster.svg",height = 3)
plot_slide(xy_clear,samples_clear,ref,col=cols_cluster)
dev.off()
```

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
save(kk_UMAP_clear,samples_clear,xy_clear,labels_clear,ref,clu,file="output/DLFPC-All-2.RData")
```



```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
subclusters=sort(names(sort(table(ref,labels_clear)[,"Layer3"],decreasing = TRUE)[1:3]))

sa_sel=(samples_clear=="151669" | samples_clear=="151670" | samples_clear=="151671" | samples_clear=="151672")

aa=apply(data_clear,2,function(x) wilcox.test(x[ref==subclusters[1] & sa_sel],x[(ref==subclusters[2] ) & sa_sel],alternative = "greater")$p.value) +
  apply(data_clear,2,function(x) wilcox.test(x[ref==subclusters[1] & sa_sel],x[(ref==subclusters[3]) & sa_sel],alternative = "greater")$p.value)
sort(aa)[1:10]
box_sel=(samples_clear=="151669" | samples_clear=="151670" | samples_clear=="151671" | samples_clear=="151672") & (ref==subclusters[1] | ref==subclusters[2] | ref==subclusters[3])
tapply(data_clear[box_sel,names(sort(aa)[1])],as.numeric(as.vector(ref[box_sel])),mean)

df=data.frame(variable=data_clear[box_sel,names(sort(aa)[1])],labels=as.numeric(as.vector(ref[box_sel])))

my_comparisons=list()
my_comparisons[[1]]=c(1,3)
my_comparisons[[2]]=c(1,2)

Nplot1=ggboxplot(df, x = "labels", y = "variable",fill="labels",
                 width = 0.8,
                 palette=cols_cluster)+  
  ylab(names(sort(aa)[1]))+ xlab("")+
  stat_compare_means(comparisons = my_comparisons,method="wilcox.test")

Nplot1



a1=apply(data_clear,2,function(x) wilcox.test(x[ref==subclusters[2] & sa_sel],x[(ref==subclusters[3]) & sa_sel],alternative = "greater")$p.value)
  a2=apply(data_clear,2,function(x) wilcox.test(x[ref==subclusters[2] & sa_sel],x[(ref==subclusters[1] ) & sa_sel],alternative = "greater")$p.value)
  
aa=a1+a2
  
sort(aa)[1:10]
a_choice=names(sort(aa)[1])
tapply(data_clear[box_sel,a_choice],as.numeric(as.vector(ref[box_sel])),mean)

df=data.frame(variable=data_clear[box_sel,a_choice],labels=as.numeric(as.vector(ref[box_sel])))

my_comparisons=list()
my_comparisons[[1]]=c(2,3)
my_comparisons[[2]]=c(1,2)

Nplot2=ggboxplot(df, x = "labels", y = "variable",fill="labels",
                 width = 0.8,
                 palette=cols_cluster)+  
  ylab(a_choice)+ xlab("")+
  stat_compare_means(comparisons = my_comparisons,method="wilcox.test")

Nplot2

aa=apply(data_clear,2,function(x) wilcox.test(x[ref==subclusters[3] & sa_sel],x[(ref==subclusters[1] ) & sa_sel],alternative = "greater")$p.value) +
  apply(data_clear,2,function(x) wilcox.test(x[ref==subclusters[3] & sa_sel],x[(ref==subclusters[2]) & sa_sel],alternative = "greater")$p.value)
sort(aa)[1:10]
tapply(data_clear[box_sel,names(sort(aa)[1])],as.numeric(as.vector(ref[box_sel])),mean)


df=data.frame(variable=data_clear[box_sel,names(sort(aa)[1])],labels=as.numeric(as.vector(ref[box_sel])))

my_comparisons=list()
my_comparisons[[1]]=c(2,3)
my_comparisons[[2]]=c(1,3)

Nplot3=ggboxplot(df, x = "labels", y = "variable",fill="labels",
                 width = 0.8,
                 palette=cols_cluster)+  
  ylab(names(sort(aa)[1]))+ xlab("")+
  stat_compare_means(comparisons = my_comparisons,method="wilcox.test")

Nplot3




library(ggalluvial)

sel_sub=subjects_clear=="Br5292"
al=data.frame(expand.grid(list(levels(labels_clear),1:7)),freq=as.numeric(table(labels_clear[sel_sub],ref[sel_sub])))
al$Var2=as.factor(al$Var2)


ggplot(data = al,
       aes(axis1 = Var1, axis2 = Var2, y = freq)) +
  geom_alluvium(aes(fill = Var2)) +
  geom_stratum() +
  geom_text(stat = "stratum",
            aes(label = after_stat(stratum))) +
  theme_void()


sel_sub=subjects_clear=="Br5595"
al=data.frame(expand.grid(list(levels(labels_clear),1:7)),freq=as.numeric(table(labels_clear[sel_sub],ref[sel_sub])))
al$Var2=as.factor(al$Var2)


ggplot(data = al,
       aes(axis1 = Var1, axis2 = Var2, y = freq)) +
  geom_alluvium(aes(fill = Var2)) +
  geom_stratum() +
  geom_text(stat = "stratum",
            aes(label = after_stat(stratum))) +
  theme_void()



sel_sub=subjects_clear=="Br8100"
al=data.frame(expand.grid(list(levels(labels_clear),1:7)),freq=as.numeric(table(labels_clear[sel_sub],ref[sel_sub])))
al$Var2=as.factor(al$Var2)


ggplot(data = al,
       aes(axis1 = Var1, axis2 = Var2, y = freq)) +
  geom_alluvium(aes(fill = Var2)) +
  geom_stratum() +
  geom_text(stat = "stratum",
            aes(label = after_stat(stratum))) +
  theme_void()











 library("GSVA")
 library("GSA")
 library("VAM")
 geneset=GSA.read.gmt("../Genesets/msigdb_v2023.2.Hs_GMTs/h.all.v2023.2.Hs.symbols.gmt")
 names(geneset$genesets)=geneset$geneset.names
 genesets=geneset$genesets

countdata <- as.matrix(t(logcounts(spe)))

# library("gprofiler2")
# genes=gconvert(rownames(spe),organism="hsapiens",target="GENECARDS",filter_na = F)$target

genes=rowData(spe)[,"gene_name"]
spot_name=colnames(spe)
colnames(countdata)=genes



li=lapply(genesets,function(x) which(genes %in% x))

VAM=vamForCollection(gene.expr=countdata, gene.set.collection=li)
pathway=VAM$distance.sq
  pathway_clear=pathway[-sel_rem,]







aa=apply(pathway_clear,2,function(x) wilcox.test(x[ref==subclusters[1] & sa_sel],x[(ref==subclusters[2] ) & sa_sel],alternative = "greater")$p.value) +
  apply(pathway_clear,2,function(x) wilcox.test(x[ref==subclusters[1] & sa_sel],x[(ref==subclusters[3]) & sa_sel],alternative = "greater")$p.value)
sort(aa)[1:10]

box_sel=(samples_clear=="151669" | samples_clear=="151670" | samples_clear=="151671" | samples_clear=="151672") & (ref==subclusters[1] | ref==subclusters[2] | ref==subclusters[3])
tapply(pathway_clear[box_sel,names(sort(aa)[1])],as.numeric(as.vector(ref[box_sel])),mean)

df=data.frame(variable=log(1+pathway_clear[box_sel,names(sort(aa)[1])]),labels=as.numeric(as.vector(ref[box_sel])))

my_comparisons=list()
my_comparisons[[1]]=c(1,3)
my_comparisons[[2]]=c(1,2)

Nplot1=ggboxplot(df, x = "labels", y = "variable",fill="labels",
                 width = 0.8,
                 palette=cols_cluster)+  
  ylab(names(sort(aa)[1]))+ xlab("")+
  stat_compare_means(comparisons = my_comparisons,method="wilcox.test")

Nplot1



a1=apply(pathway_clear,2,function(x) wilcox.test(x[ref==subclusters[2] & sa_sel],x[(ref==subclusters[3]) & sa_sel],alternative = "greater")$p.value)
  a2=apply(pathway_clear,2,function(x) wilcox.test(x[ref==subclusters[2] & sa_sel],x[(ref==subclusters[1] ) & sa_sel],alternative = "greater")$p.value)
  
aa=a1+a2
  
sort(aa)[1:10]
a_choice=names(sort(aa)[1])
tapply(pathway_clear[box_sel,a_choice],as.numeric(as.vector(ref[box_sel])),mean)

df=data.frame(variable=pathway_clear[box_sel,a_choice],labels=as.numeric(as.vector(ref[box_sel])))

my_comparisons=list()
my_comparisons[[1]]=c(2,3)
my_comparisons[[2]]=c(1,2)

Nplot2=ggboxplot(df, x = "labels", y = "variable",fill="labels",
                 width = 0.8,
                 palette=cols_cluster)+  
  ylab(a_choice)+ xlab("")+
  stat_compare_means(comparisons = my_comparisons,method="wilcox.test")

Nplot2

aa=apply(pathway_clear,2,function(x) wilcox.test(x[ref==subclusters[3] & sa_sel],x[(ref==subclusters[1] ) & sa_sel],alternative = "greater")$p.value) +
  apply(pathway_clear,2,function(x) wilcox.test(x[ref==subclusters[3] & sa_sel],x[(ref==subclusters[2]) & sa_sel],alternative = "greater")$p.value)
sort(aa)[1:10]
tapply(pathway_clear[box_sel,names(sort(aa)[1])],as.numeric(as.vector(ref[box_sel])),mean)


df=data.frame(variable=pathway_clear[box_sel,names(sort(aa)[1])],labels=as.numeric(as.vector(ref[box_sel])))

my_comparisons=list()
my_comparisons[[1]]=c(2,3)
my_comparisons[[2]]=c(1,3)

Nplot3=ggboxplot(df, x = "labels", y = "variable",fill="labels",
                 width = 0.8,
                 palette=cols_cluster)+  
  ylab(names(sort(aa)[1]))+ xlab("")+
  stat_compare_means(comparisons = my_comparisons,method="wilcox.test")

Nplot3



```

TRAJECTORY

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}


d <- slingshot(kk_UMAP_clear, clusterLabels = clu)
trajectory=d@metadata$curves$Lineage1$s
k=Rnanoflann::nn(trajectory,kk_UMAP_clear,1)
map_color=rainbow(nrow(trajectory))[k$indices]
plot(kk_UMAP_clear,pch=20,col=map_color)

plot_slide(xy_clear,samples_clear,k$indices,col=rainbow(nrow(trajectory)))

```

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}
#source("code/DLPFC_comparison.R")
```







