---
title: "Installation"
output:
  workflowr::wflow_html:
    toc: True
editor_options:
  chunk_output_type: console
---

# Introduction

Welcome to the KODAMA software installation tutorial page. This guide will walk you through the steps to install and set up KODAMA, a powerful tool for metabolomics data analysis.

## Installation Steps

### Step 1: Prerequisites

Before installing KODAMA, ensure you have the following:

- **R (>= 4.0.3)**: Make sure you have R installed on your system. You can download it from [CRAN](https://cran.r-project.org/).

- **RStudio (Optional)**: While not required, RStudio provides a convenient interface for working with R projects. You can download it from [RStudio's official site](https://www.rstudio.com/products/rstudio/download/).


### Step 2: Install KODAMA Package

Use the following commands to install KODAMA from GitHub using the `devtools` package:

```{r,fig.width=10,fig.height=4,echo=TRUE,eval=FALSE,warning=FALSE,message=FALSE}
if (!require(devtools)) {
  install.packages("devtools")
}
devtools::install_github("tkcaccia/KODAMA")
devtools::install_github("tkcaccia/KODAMAextra")
```

### Step 3: Load KODAMA Package

Once installed, load the KODAMA package into your R session:

```{r,fig.width=10,fig.height=4,echo=TRUE,eval=FALSE,warning=FALSE,message=FALSE}
library(KODAMA)
```

### Step 5: Verify Installation

To verify that KODAMA has been installed correctly, you can check its version:

```{r,fig.width=10,fig.height=4,echo=TRUE,eval=FALSE,warning=FALSE,message=FALSE}
packageVersion("KODAMA")
```

## Support

For any issues or questions related to KODAMA, please [open an issue](https://github.com/tkcaccia/KODAMA/issues) on the GitHub repository.
