---
title: "About"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---


Describe your project.
The data can be downloaded using the following script: [VisiumHD_CRC_download.sh](code/VisiumHD_CRC_download.sh). This script facilitates access to the raw data, which will then be preprocessed and analyzed in the subsequent steps of our pipeline.

The data can be downloaded using the following script: [VisiumHD_CRC_download.sh](https://github.com/tkcaccia/KODAMA-Analysis/blob/main/code/VisiumHD_CRC_download.sh). This script facilitates access to the raw data, which will then be preprocessed and analyzed in the subsequent steps of our pipeline.


```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
library("ggplot2")
library("patchwork")
library("dplyr")
library("Seurat")
library("KODAMA")
library("KODAMAextra")
library("bigmemory")

localdir="../Colorectal/outs/"
object <- Load10X_Spatial(data.dir = localdir, bin.size = c(8))


 vln.plot <- VlnPlot(object, features = "nCount_Spatial.008um", pt.size = 0) + NoLegend()
 count.plot <- SpatialFeaturePlot(object, features = "nCount_Spatial.008um", pt.size.factor = 1.2) +
   theme(legend.position = "right")

nCount_Spatial=colSums(object@assays$Spatial.008um$counts)
#w= which(nCount_Spatial >10)
#object@assays$Spatial.008um$counts= object@assays$Spatial.008um$counts[,w]
#object@meta.data=object@meta.data[w,]

sp_obj <- subset(
  object,
  subset = nCount_Spatial.008um > 100)



nCount_Spatial=colSums(sp_obj@assays$Spatial.008um$counts)


 counts=sp_obj@assays$Spatial.008um$counts
 is_mito <- grepl("(^MT-)|(^mt-)", rownames(counts))
 counts <- counts[!is_mito,]

 filter_genes_ncounts=1
 filter_genes_pcspots=0.5
 nspots <- ceiling(filter_genes_pcspots/100 *  ncol(counts))
 ix_remove <- rowSums(counts >= filter_genes_ncounts) <   nspots
 counts <- counts[!ix_remove,]

 QCgenes <- rownames(counts)

 VariableFeatures(sp_obj) = QCgenes

 rm(counts)


DefaultAssay(sp_obj) <- "Spatial.008um"
sp_obj <- NormalizeData(sp_obj)


sp_obj <- FindVariableFeatures(sp_obj)
sp_obj <- ScaleData(sp_obj)

xy=as.matrix(GetTissueCoordinates(sp_obj)[,1:2])

sp_obj <- RunPCA(sp_obj, reduction.name = "pca.008um")

dim(sp_obj)

plot(Seurat::Embeddings(sp_obj, reduction = "pca.008um"))
```

```{r, fig.width=10, fig.height=4}

sp_obj <- RunKODAMAmatrix(sp_obj, reduction = "pca.008um",
                          FUN= "PLS" ,
                          landmarks = 10000,
                          splitting = 100,
                          f.par.pls = 50,
                          spatial.resolution = 0.4,
                          n.cores=8)

  print("KODAMA finished")
  
     config=umap.defaults
     config$n_threads = 8
     config$n_sgd_threads = "auto"
     sp_obj <- RunKODAMAvisualization(sp_obj, method = "UMAP",config=config)

     
kk_UMAP=Seurat::Embeddings(sp_obj, reduction = "KODAMA")

     

rr=read.csv("data/spots_classification_ALL.csv",sep=",")
ss=strsplit(rr[,2],":")
ss=unlist(lapply(ss, function(x) x[2]))
ss=strsplit(ss,",")
ss=unlist(lapply(ss, function(x) x[1]))
ss=gsub("\"","",ss)

rr[,2]=ss
n=ave(1:length(rr[,1]), rr[,1], FUN = seq_along)
rr=rr[n==1,]
rownames(rr)=rr[,1]
rr=rr[rownames(kk_UMAP),]

table(rr[,"classification"])

library(ggplot2)
cols <- c("#669bbc", "#81b291", "#f2cc8f", "#adc178",
          "#dde5b6", "#a8dadc", "#e5989b", "#e07a5f",
          "#ffe5b6", "#ffdadc", "#e5ff9b", "#ffff11")

labels=as.factor(rr[,"classification"])
par(xpd = T, mar = par()$mar + c(0,0,0,7))

plot(kk_UMAP,cex=0.5,pch=20,col=cols[labels])
legend(max(kk_UMAP[,1])+0.05*dist(range(kk_UMAP[,1])), max(kk_UMAP[,2]),
       levels(labels),
       col = cols,
       cex = 0.8,
       pch=20)



```


```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE, echo=FALSE}

save(kk_UMAP,xy,file="output/VisiumHD3.RData")
```


