---
title: "About"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---


Describe your project.
The data can be downloaded using the following script: [VisiumHD_CRC_download.sh](code/VisiumHD_CRC_download.sh). This script facilitates access to the raw data, which will then be preprocessed and analyzed in the subsequent steps of our pipeline.

The data can be downloaded using the following script: [VisiumHD_CRC_download.sh](https://github.com/tkcaccia/KODAMA-Analysis/blob/main/code/VisiumHD_CRC_download.sh). This script facilitates access to the raw data, which will then be preprocessed and analyzed in the subsequent steps of our pipeline.


```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
library("ggplot2")
library("patchwork")
library("dplyr")
library("Seurat")
library("KODAMA")
library("KODAMAextra")
library("bigmemory")

localdir="../Colorectal/outs/"
object <- Load10X_Spatial(data.dir = localdir, bin.size = c(8))

image=as.raster(object@images$slice1.008um@image)
save(image,file="output/CRC-image.RData")
#object@images$slice1.008um@scale.factors$hires

# plot(image,xlim=c(320,530),ylim=c(200,410))
# points(xy[,2]*0.007973422,nrow(image)-xy[,1]*0.007973422,pch=20)

#xy=as.matrix(GetTissueCoordinates(sp_obj)[,1:2])

#image=as.raster(imgData(object)$data[[1]])

#xy_sel=spatialCoords(spe_sub)
#xy_sel=xy_sel*scaleFactors(spe_sub)
#xy_sel[,2]=nrow(image)-xy_sel[,2]


 vln.plot <- VlnPlot(object, features = "nCount_Spatial.008um", pt.size = 0) + NoLegend()
 count.plot <- SpatialFeaturePlot(object, features = "nCount_Spatial.008um", pt.size.factor = 1.2) +
   theme(legend.position = "right")

nCount_Spatial=colSums(object@assays$Spatial.008um$counts)
#w= which(nCount_Spatial >10)
#object@assays$Spatial.008um$counts= object@assays$Spatial.008um$counts[,w]
#object@meta.data=object@meta.data[w,]

sp_obj <- subset(
  object,
  subset = nCount_Spatial.008um > 100)



nCount_Spatial=colSums(sp_obj@assays$Spatial.008um$counts)


 counts=sp_obj@assays$Spatial.008um$counts
 is_mito <- grepl("(^MT-)|(^mt-)", rownames(counts))
 counts <- counts[!is_mito,]

 filter_genes_ncounts=1
 filter_genes_pcspots=0.5
 nspots <- ceiling(filter_genes_pcspots/100 *  ncol(counts))
 ix_remove <- rowSums(counts >= filter_genes_ncounts) <   nspots
 counts <- counts[!ix_remove,]

 QCgenes <- rownames(counts)

 VariableFeatures(sp_obj) = QCgenes

 rm(counts)


DefaultAssay(sp_obj) <- "Spatial.008um"
sp_obj <- NormalizeData(sp_obj)


sp_obj <- FindVariableFeatures(sp_obj)
sp_obj <- ScaleData(sp_obj)

xy=as.matrix(GetTissueCoordinates(sp_obj)[,1:2])

sp_obj <- RunPCA(sp_obj, reduction.name = "pca.008um")

dim(sp_obj)

plot(Seurat::Embeddings(sp_obj, reduction = "pca.008um"))
```

```{r, fig.width=10, fig.height=8}

#sp_obj <- RunKODAMAmatrix(sp_obj, reduction = "pca.008um",
#                          FUN= "PLS" ,
#                          landmarks = 10000,
#                          splitting = 100,
#                          f.par.pls = 50,
#                          spatial.resolution = 0.4,
#                          n.cores=8)

#  print("KODAMA finished")
  
#     config=umap.defaults
#     config$n_threads = 8
#     config$n_sgd_threads = "auto"
#     sp_obj <- RunKODAMAvisualization(sp_obj, method = "UMAP",config=config)

     
# kk_UMAP=Seurat::Embeddings(sp_obj, reduction = "KODAMA")

# save(kk_UMAP,xy,file="output/VisiumHD.RData")

load("output/VisiumHD3.RData")   

rr=read.csv("data/spots_classification_ALL.csv",sep=",")
ss=strsplit(rr[,2],":")
ss=unlist(lapply(ss, function(x) x[2]))
ss=strsplit(ss,",")
ss=unlist(lapply(ss, function(x) x[1]))
ss=gsub("\"","",ss)

rr[,2]=ss
n=ave(1:length(rr[,1]), rr[,1], FUN = seq_along)
rr=rr[n==1,]
rownames(rr)=rr[,1]
rr=rr[rownames(kk_UMAP),]

table(rr[,"classification"])



library(ggplot2)


cols=sample(rainbow(15))
labels=as.factor(rr[,"classification"])
par(xpd = T, mar = par()$mar + c(0,0,0,7))

plot(kk_UMAP,cex=0.5,pch=20,col=cols[labels])
legend(max(kk_UMAP[,1])+0.05*dist(range(kk_UMAP[,1])), max(kk_UMAP[,2]),
       levels(labels),
       col = cols,
       cex = 0.8,
       pch=20)


load("data/trajectories_VISIUMHD.RData")

data=sp_obj@assays$Spatial.008um$data[rownames(sp_obj@assays$Spatial.008um$scale.data),]
data=as.matrix(data)
data=t(data)

mm1=new_trajectory (kk_UMAP,data = data,trace=tra1$xy)
mm2=new_trajectory (kk_UMAP,data = data,trace=tra2$xy)
mm3=new_trajectory (kk_UMAP,data = data,trace=tra3$xy)


traj=rbind(mm1$trajectory,
           mm2$trajectory,
           mm3$trajectory)
y=rep(1:20,3)

```

```{r, fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
ma=multi_analysis(traj,y,FUN="correlation.test",method="spearman")
ma=ma[order(as.numeric(ma$`p-value`)),]
colnames(ma)=c("Feature   ","rho   ","p-value   ","FDR   ")
```

```{r, fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
knitr::kable(ma[1:30,],row.names=FALSE)
```

```{r, fig.width=10, fig.height=8, warning=FALSE, message=FALSE,results = "hide"}


lab=rr[,"classification"]
sel=lab==" stroma invasive carcinoma" | lab==" stroma intermediate"
data.sel=data[which(sel),]
data.sel=data.sel[,colSums(data.sel)>0]
lab=as.factor(as.vector(lab[which(sel)]))

ma=multi_analysis(data.sel,lab,range="95%CI")

ma=ma[order(as.numeric(ma$`p-value`)),]
```


```{r, fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
knitr::kable(ma[1:30,],row.names=FALSE)
```


```{r, fig.width=10, fig.height=8, warning=FALSE, message=FALSE,results = "hide"}

lab=rr[,"classification"]
sel=lab==" stroma invasive carcinoma" | lab==" desmoplastic submucosa"
data.sel=data[which(sel),]
data.sel=data.sel[,colSums(data.sel)>0]

lab=as.factor(as.vector(lab[which(sel)]))

ma=multi_analysis(data.sel,lab,range="95%CI")

ma=ma[order(as.numeric(ma$`p-value`)),]
```


```{r, fig.width=10, fig.height=8, warning=FALSE, message=FALSE}
knitr::kable(ma[1:30,],row.names=FALSE)
```



